---
title: "Beautiful graphics with ggplot2"
---

This document is the web-based version of a presentation given through the University of Idaho library workshop series on September 12, 2017. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, message = FALSE, eval = TRUE, warning = FALSE)

#Load libraries:

#Data organizing:
library(tidyverse)
library(data.table)

#Palettes and visualization:
library(viridis)
library(RColorBrewer)
library(wesanderson)
library(ggthemes)
library(ggjoy)

#Data
library(gapminder)
library(maptools)

#Presentation tools:
library(knitr)
library(revealjs)
```


## What is ggplot2?

-  an R package for data visualization

- implements the "grammar of graphics"


## Why ggplot2?

- Popular (well-supported, great community)

- Open source (like all of R)

- Easy to use (after a learning curve)

- Aesthetically pleasing

- Built for multi-variate data

- Reproducible figures

## Goals for today

Teach you enough that you know how to teach yourself more!

- Introduce "grammar of graphics" concepts

- Practice!
      + samples with built-in data
      + preparing data
      + samples with (more interesting?) data


## Grammar of graphics

- data with variables mapped to aesthetics

- one or more geometric layers 

- a scale for each aesthetic mapping

- a coordinate system

- a facet specification

If this is confusing, don't worry - the main message here is that there is a structured grammar underlying ggplot2. 

## Let's try it!

First, we need data. 

Let's use the built-in R dataset, "diamonds".

```{r diamonds, echo = FALSE}
kable(head(diamonds))
```

## Data

```{r diamonds1, echo = TRUE}
ggplot(data = diamonds)

```

## Aesthetics: aes()

```{r diamonds2, echo = TRUE}
ggplot(data = diamonds,
       aes(x = carat, y = price))

```

## Geometric object: geom_

```{r diamonds3, echo = TRUE}
ggplot(data = diamonds, 
       aes(x = carat, y = price)) +
  geom_point()

```

## Add more aesthetics...

```{r diamonds4, echo = TRUE}
ggplot(data = diamonds, 
       aes(x = carat, y = price, color = clarity)) +
  geom_point()

```


## Why didn't we do this?

```{r diamonds5, echo = TRUE}
ggplot(data = diamonds, 
       aes(x = carat, y = price), color = clarity) +
  geom_point()

```

Putting `rcolor = clarity` outside of the "aesthetics" call meant that ggplot didn't know that it was supposed to map data to the "color" aesthetic. 

## We could get this instead:

```{r diamonds6, echo = TRUE}
ggplot(data = diamonds, 
       aes(x = carat, y = price)) +
  geom_point(color = "magenta")

```

## Our original plot
(with a small difference:)

```{r diamonds7, echo = TRUE}
ggplot(data = diamonds, 
       aes(x = carat, y = price)) +
  geom_point(aes(color = clarity))

```

## Add more geoms:

```{r diamonds8, echo = TRUE, message = FALSE}
ggplot(data = diamonds, 
       aes(x = carat, y = price)) +
  geom_point(aes(color = clarity)) +
  geom_smooth()
```

## Change geom appearance:

```{r diamonds9, echo = TRUE, message = FALSE}
ggplot(data = diamonds, 
       aes(x = carat, y = price)) +
  geom_point(aes(color = clarity)) +
  geom_smooth(color = "black", size = 0.8, linetype = 2)

```

## Make it look nicer:

```{r diamonds12, echo = TRUE, message = FALSE}
ggplot(data = diamonds, 
       aes(x = carat, y = price)) +
  geom_point(aes(color = clarity)) +
  geom_smooth(color = "black", size = 0.8, linetype = 2) +
  theme_few()
```


## Try facets!

```{r diamonds10, echo = TRUE, message = FALSE}
ggplot(data = diamonds, 
       aes(x = carat, y = price)) +
  geom_point(aes(color = clarity)) +
  geom_smooth(color = "black", size = 0.8, linetype = 2) +
  facet_wrap(~cut)

```

## More facets!

```{r diamonds11, echo = TRUE, message = FALSE}
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = clarity), size = 0.5) +
  facet_grid(color~cut)
```

## Try out color scales

ColorBrewer is useful and popular:
```{r set plot}
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = clarity)) +
  theme_few() + 
  scale_color_brewer(type = "qual", palette = "Set2")

```

## Different plot types?

```{r diamonds13, echo = TRUE, message = FALSE}
ggplot(data = diamonds, 
       aes(x = clarity, y = price)) +
  geom_violin() 

```

## Try an extension:

```{r diamonds14, echo = TRUE, message = FALSE}
ggplot(data = diamonds, 
       aes(x = price, y = cut)) +
  geom_joy()

```

## Adjust color scales:

```{r diamonds viridis, echo = TRUE, message = FALSE}
ggplot(data = diamonds, 
       aes(x = price, y = cut, color = cut, fill = cut)) +
  geom_joy(alpha = 0.6, scale = 5) +
  scale_fill_viridis(option = "A", discrete = TRUE) +
  scale_color_viridis(option = "A", discrete = TRUE) + 
  theme_few()

```

## Another color scale: 
```{r diamonds wes, echo = TRUE, message = FALSE}
ggplot(data = diamonds, 
       aes(x = price, y = cut, fill = cut)) +
  geom_joy(alpha = 1, scale = 5) +
  scale_fill_manual(values = wes_palette("Darjeeling")) +
  theme_few()

```

## Sometimes we need to transform data.

```{r iris melt}
kable(head(iris))

```

## Wide or long?

```{r iris melt2}
iris_long <- melt(iris, id.vars = ("Species"))
kable(head(iris_long))

```

## Plot the data:

```{r iris plot}
ggplot(iris_long, aes(x = Species, y = value, fill = variable)) +
  geom_bar(stat = 'identity', width = 1) +
  theme_bw()

```

## Change coordinates

```{r iris plot polar}
ggplot(iris_long, 
       aes(x = Species, y = value, color = variable, fill = variable)) +
  geom_bar(stat = 'identity', width = 1) +
  coord_polar(theta = 'x') +
  theme_bw()

```


## Try out a new dataset:

```{r look at gapminder}
kable(head(gapminder))
```

## Take a look at the data:

```{r test gapminder}
ggplot(gapminder,
       aes(x = year, y = lifeExp, color = gdpPercap)) +
  geom_line(aes(group = country)) +
  facet_grid(continent~.) +
  scale_color_viridis(trans = "log") +
  theme_few()

```

## A scatter plot:

```{r gapminder plots}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point()

```

## Who's the outlier?

```{r gapminder outlier}
gapminder %>% filter(gdpPercap > 60000)

```


## Deal with overplotting:

```{r gapminder hex}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_hex()

```

## Another way to avoid overplotting:

```{r gapminder contour}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_density2d(aes(color = ..level..), bins = 20) +
  scale_color_viridis()

```


## A little more complicated:

```{r make gapminder faceted}
p <- ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(color = continent, size = pop), alpha = 0.8) +
  scale_x_continuous(trans = 'log') +
  facet_wrap(~year) +
  scale_color_brewer(type = "Qual", palette = "Accent") +
  theme_hc(bgc = 'darkunica') +
  theme(text = element_text(size = 9))

```

## A little more complicated:

```{r view gapminder faceted, fig.height = 5, fig.width = 9}
p

```

## Let's make a map!

Prepare data: 
```{r gapminder map}
country_df <- map_data('world') %>%
  rename("country" = "region") 
country_df$country[country_df$country == "USA"] <- "United States"

#Take the mean across all years for each country:
gapminder_means <- gapminder %>% 
  group_by(country, continent) %>%
  summarise(lifeExp = mean(lifeExp),
            pop = mean(pop),
            gdpPercap = mean(gdpPercap))

plot_dat <- left_join(gapminder_means, country_df, by = "country")
```

## Make the map: 

```{r gapminder map2}
ggplot(plot_dat) +
  geom_polygon(aes(x = long, y = lat, fill = lifeExp, group = group)) +
  scale_fill_viridis(option = "A") + 
  coord_quickmap() +
  theme_few()
```


## More plot ideas?

- What questions could we ask with this data?

- How could we visually answer those questions?

## Where to learn more:
<small>

- Stack Overflow/Google
- Books:
    + R for Data Science: http://r4ds.had.co.nz/program-intro.html
    + Ggplot2: Elegant graphics for data analysis: https://www.amazon.com/ggplot2-Elegant-Graphics-Data-Analysis/dp/0387981403 (may also be available for free at github).
    + Data visualization for social science: http://socviz.co/
- Papers:
    + Wickham grammar of graphics: http://byrneslab.net/classes/biol607/readings/wickham_layered-grammar.pdf
    + Wickham tidy data: https://www.jstatsoft.org/article/view/v059i10
- Tutorials, cheatsheets, websites:
    + https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf
    + http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html
    + http://www.ggplot2-exts.org/gallery/
- Twitter:
    + @hadleywickham, @ClausWilke, @JennyBryan, @RLadiesGlobal, @rstudiotips
    
</small>



## Questions?
Thanks!

<small> Adrienne Marshall
mars7850@vandals.uidaho.edu
</small>






